Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2016-10-17 article.tex
Kommentare von Jens

%% an abstract and keywords
\Abstract{
In a wide variety of research fields, dynamic modeling is employed as an instrument to learn and understand complex systems. The differential equations involved in this process are mostly non-linear and depend on many parameters whose values decide upon the characteristics of the emergent system. The inverse problem, i.e.~the inference or estimation of parameter values from observed data, is interesting from two points of view. First, the existence point of view, dealing with the question whether the system is able to  produce the observed dynamics for any parameter values. Second, the identifiability point of view, investigating invariance of the prediction under change of parameter values, as well as the quantification of parameter uncertainty.

@@

The key elements of the framework and the methodology implemented in \pkg{dMod} are highlighted by an application on a three-compartment transporter model within this paper.

@@

}
\Keywords{Dynamic models, parameter estimation, code generation, maximum likelihood}
\Plainkeywords{Dynamic models, parameter estimation, code generation, maximum likelihood} %% without formatting
%% at least one keyword must be supplied

@@

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
\section{Introduction}
Dynamic models are found in several research fields, such as physics, biology or finance. In all these fields, models link theoretical concepts and empirical evidence.
%In a dynamic model, the single terms correspond to mechanisms or processes which are combined to a whole, giving rise to a complex system. If the single terms are associated to parameters, parameter estimation can identify those processes which are crucial to explain the observation.
Single mechanisms or single processes of a complex system are represented by single terms occurring in the equations of a dynamic model. Parameter estimation can then identify those processes which are crucial to explain the observation.
In that sense, parameter estimation can be employed as an instrument to \textit{understand} complex systems. Once, the link between observation and model is established by the estimated parameters, questions about their identifiability arise. The parameter space needs to be explored to analyze whether the estimate is unique and to determine confidence bounds.

@@

Another distinguishing feature of \pkg{dMod} is the handling of non-identifiability of parameters, a phenomenon occurring frequently in the context of parameter estimation in dynamic systems.
In some cases, non-identifiability has structural reasons. The differential equations bear certain symmetries and these can or cannot be broken by observation functions. A functional relationship between parameters that leaves the observation invariant is the consequence of the latter. In other cases, the data allows to determine a unique optimum but other solutions, although worse, cannot be statistically rejected. The \pkg{dMod} package deals with parameter identifiability and parameter uncertainty by the profile-likelihood method \citep{murphy2000profile, kreutz2013profile}. The method has proven especially useful in the case of non-identifiable parameters where results obtained from both, the quadratic approximation by the variance-covariance matrix and MCMC sampling can be misleading \citep{raue2013joining}. Besides parameter uncertainties, the profile likelihood method allows to estimate prediction uncertaint \citep{kreutz2012likelihood}. It therefore supports the planning of new informative experiments or suggests possible model reductions.

@@

\subsection{Dynamic models and model sensitivities}
Dynamic models describe systems with states $x$, usually quantifying the involved species, their interaction and evolution over timer. %  as e.g. given by chemical reaction networks.
The time evolution of the states is expressed via a set of ODEs, $\dot x = f(x)$. Although constituting quite a special class of dynamic systems, chemical reaction networks formulated by the law of mass action as considered here allow for surprisingly general applications. Typical ODE examples derived from the law of mass action are:
\begin{itemize}
	\item $\dot x_A = k_p$, constant production of $x_A$ ($\emptyset \rightarrow A$)
	\item $\dot x_C = k_c \cdot x_A \cdot x_B = - \dot x_A = - \dot x_B$, complex formation ($A+B \rightarrow C$)
	\item $\dot x_C = - k_d \cdot x_C$, proportional degradation of $x_C$ ($C \rightarrow \emptyset$)
\end{itemize}
each generally written as
\begin{align}
	\dot x = \underbrace{k_0 \left(\prod x_i^{\nu_i}\right)}_{v_0(x, k_0)} (\nu' - \nu) \quad\leftrightarrow\quad (\nu_1 X_1 + \dots + \nu_n X_n \longrightarrow \nu'_1 X_1 + \dots + \nu'_n X_n)
	\label{eq:ma}
\end{align}
where $X_1, \dots, X_n$ are the $n$ species involved, $\nu$ and $\nu'$ encode the stoichiometry of the reaction and $k_0$ is the reaction rate. When several reactions are involved, eq.~\eqref{eq:ma} generalizes to
\begin{align}
	\dot x = S\cdot v(x, k)
	\label{eq:Sv}
\end{align}
%$\dot x_l = \sum_{j = 1\dots m}  k_{j} \prod_{i = 1\dots n} \nu_{i,j} x_i^{\nu_{i,j}}$ for a system of $n$ states and $m$ reactions with rate constants $p = \{ k_j, j = 1\dots m\}$ and stoichiometric constants $\nu_{i,j} \in \mathbb{Z}$.
%An even more general expression of the system is given by $\dot x = S v(x,p)$
with the stoichiometry matrix $S$
%containing the stoichiometric constants $\nu_{i,j}$
obtained from the collection of vectors $\nu' - \nu$ for reactions $1, \dots, r$,
and the flux vector $v(x,k)\in\mathbb R^{r}$.
Eq.~\eqref{eq:Sv} also holds for general ODEs that do not follow from mass action kinetics. Moreover the system may be explicitly time-dependent and may contain forcings $u(t)$ which e.g.~describe the external stimulation of the system. A general form of the dynamic model is therefore given by
\begin{align}
	\dot x = f(x,k,u,t) = S \cdot v(x,k,u,t), \quad\quad \textrm{with }  x(t = 0) = x_0.
\end{align}
The solution $x(t,\theta)$ for a given parameter vector $\theta = (k, x_0)$ is called model prediction. %The parameters $\theta$ contain the rate constants, initial values $x_0$ and additional parameters included in the forcings.
It is generally assumed that the stoichiometry matrix $S$ of the system is known.
Besides the model predition itself, also the sensitivity of the prediction to changes in the parameter values is of interest.
The sensitivities $s_i = \frac{\partial x}{\partial \theta_i}$ satisfy the \textit{sensitivity equations}
\begin{align}
	\dot s_i  = \frac{\partial f}{\partial x} s_i + \frac{\partial f}{\partial \theta_i}, \qquad \textrm{with }s_i(t = 0) = \frac{\partial x_0}{\partial \theta_i},
\end{align}
a system of ordinary differential equations that in general depend on the states $x$ and, therefore, need to be solved jointly with the original ODE.

Often the experimentally observed quantities $y$ do not directly correspond to the species $x$ described by the ODE, but are obtained via an observation function $g: \mathbb{R}^n \rightarrow \mathbb{R}^m$,  \begin{align}
	y = g(x, c)
	\label{eq:observation}
\end{align}
with the observation parameters $c$. Examples for observation functions are scaling and offset transformations, $y = c_s \cdot x + c_o$, or the measurement of a superposition of species, $y_i = \sum_j c_j x_j$. The model prediction for the observed states is obtained by evaluating the observation function on the solution of the ODE. Following the chain rule of differentation also the sensitivities of the observed states $\frac{\partial y}{\partial \theta_i} = \frac{\partial g}{\partial x} \frac{\partial x}{\partial\theta_i}  + \frac{\partial g}{\partial \theta_i}$ are obtained. Here, the parameter vector $\theta$ has been augmented by the observation parameters, $\theta = (k, x_0, c)$.

@@

The easiest and one of the most frequent situations is when $\phi$ follows a normal distribution, $\phi\sim\mathcal N(\mu(\theta), \Sigma)$. When measurements are statistically independent, the variance-covariance matrix $\Sigma = {\rm diag}(\sigma_1^2, \dots, \sigma_N^2)$ is diagonal and $\phi$ factorizes. The Likelihood function becomes
\begin{align}
	L(\theta) &= \prod_{i = 1}^N \frac{1}{\sqrt{2\pi\sigma_i^2}} e^{-\frac{\mu_i(\theta)^2}{2\sigma_i^2}} .
	\label{eq:likelihood}
\end{align}
The connection between the statistical model and the observed data is given by $\mu_i(\theta) = y_i(\theta) - y_i^D$. Taking twice the negative logarithm, eq.~\eqref{eq:likelihood} turns into
\begin{align}
	l(\theta) = -2\log L(\theta) &= \sum_i \left[\left(\frac{y_i(\theta) - y_i^D}{\sigma_i}\right)^2 + \log(2\pi\sigma_i^2)\right],
	\label{eq:nll}
\end{align}
thus converting the maximization of $L(\theta)$ to a minimization of $l(\theta)$.
Assuming that the data uncertainties $\sigma_i$ are known, eq.~\eqref{eq:nll} is the weighted least-squares function shifted by a constant. For unknown $\sigma_i$, eq.~\eqref{eq:nll} can be optimized with respect to $\theta$ and $\sigma$ jointly, yielding the MLE $\hat\theta^{\prime} = (\hat\theta, \hat\sigma)$.

@@

Optimization by the Newton method attempts to iteratively find the root of the gradient $\nabla l (\theta)$ by the recursion
\begin{align}
	\theta^{(n+1)} = \theta^{(n)} - Hl(\theta)^{-1}\nabla l(\theta),
	\label{eq:newton}
\end{align}
where $Hl(\theta)$ denotes the Hessian, i.e.~the matrix of second derivatives of $l(\theta)$. The fact that for normally distributed noise the log-likelihood is a least squares function, has a big advantage: gradient and (approximate) Hessian can be computed from first-order derivatives,
\begin{align}
	\nabla l(\theta) = 2 r(\theta) J(\theta) \quad\quad Hl(\theta) \approx 2 J(\theta)^T J(\theta),
	\label{eq:derivatives}
\end{align}
with the weighted residual vector $r_i(\theta) = \frac{y_i(\theta) - y_i^D}{\sigma_i}$ and its first derivative, the Jacobian matrix
\begin{align}
	J_{ij}(\theta) = \frac{\partial r_i}{\partial \theta_j}(\theta) = \frac{1}{\sigma_i} \frac{\partial y_i}{\partial \theta_j}(\theta). \label{eq:jacobian}
\end{align}
As indicated by eqs.~\eqref{eq:derivatives}-\eqref{eq:jacobian},  first order model sensitivities $\frac{\partial y_i}{\partial\theta_j}$ are sufficient to compute gradient and Hessian in good approximation.

@@

Parameter estimates $\hat \theta$  are obtained by non-linear optimization of the log-likelihood function $l(\theta)$. Although being a function of the parameters,  the log-likelihood depends on the data, too. Consequently, "equivalent" random realizations of the data would lead to different parameter estimates $\hat\theta$. Given $\hat\theta$ for one random realization of the data, the question is, if we can reject other parameters $\theta$ with a certain confidence level. This leads to the related question of parameter convidence intervals.

A useful tool to derive confidence intervals is the profile likelihood. Consider a parameter of interest, $\theta_i$, being one of the parameters in the vector $\theta$. Furthermore, let $c_{\tau}(\theta) = \theta_i - \tau = 0$ be a parameter constraint.
Then, extending the likelihood with the constraint via a Lagrange multiplier $\lambda$ yields
\begin{align}
	\begin{aligned}
		{\rm pl}_i:\quad \mathbb R & \longrightarrow \mathbb R\\
		\tau & \longmapsto \min_{(\theta, \lambda)} \big[ l(\theta) + \lambda c_{\tau}(\theta) \big]
	\end{aligned}
	\label{eq:pl}
\end{align}
which called the profile likelihood of the $i^{\;\rm th}$ parameter. The path $\tau \mapsto (\hat\theta_{\tau}, \hat\lambda_{\tau})$ along the minimizing parameters is called the profile likelihood path. Hence, the profile likelihood ${\rm pl}_i$ returns the minimal log-likelihood under the constraint $\theta_i \stackrel{!}{=} \tau$. By construction, the inequality $D := {\rm pl}_i(\tau) - {\rm pl}_i(\hat \theta_i) \geq 0$ holds for all $\tau$, assuming equality at least for $\tau = \hat\theta_i$. Interpreting the unconstrained model as the null model and that with $\theta_i$ fixed to $\tau$ as the alternative model, the value $D$ is twice the log of the likelihood-ratio between those. Hence, a likelihood-ratio test can be performed to accept or reject the alternative model at a given confidence level. To compute the corresponding threshold, we assume Wilks' theorem according to which the thresholds are the quantiles of the $\chi^2$ distribution with one degree of freedom. Consequently, the $1\sigma, 2\sigma, \dots, n\sigma$-confidence intervals of $\theta_i$ are those values of $\tau$ for which ${\rm pl}_i(\tau) \leq 1, 2, \dots, n$, respectively.

@@

\pkg{dMod} makes use of derivatives wherever possible. It is one of the core functionalities of the \pkg{cOde} package upon which \pkg{dMod} is based to augment a system of ODEs by its sensitivity equations. If many parameters are involved and the system consists of many states, the number of equations quickly grows. Solving the equations can be considerably accelerated by utilizing compiled code. \pkg{dMod} provides the \code{odemodel} class. An object of class \code{odemodel} is generated from an \code{eqnlist} or \code{eqnvec}. In the background, \proglang{C} code for the ODE and the combined system of ODE and sensitivity equations is generated, written to the working directory and compiled. The \code{odemodel} object keeps track of the shared objects. It is the basis of a prediction function $x(t, p)$ generated by the \code{Xs()} command.

@@

\pkg{dMod} seeks to stay close to the mathematical formulation, i.e.~\code{x <- Xs(myodemodel)} will indeed return an \proglang{R} function, an object of class \code{prdfn}, which expects arguments \code{times} and \code{pars} and turns them into a model prediction. However, as indicated by the \code{"s"} in \code{"Xs"}, not only the solution $x(t)$ is returned, the sensitivities $\frac{\partial x}{\partial p}(t)$ are returned, too.

Besides parameters, the prediction might also depend on forcings $u(t)$ and sudden events, e.g.~setting states to 0 at a predefined point in time. Since forcings and events are by default not variable, they are defined for a prediction function the moment it is generated, \code{x <- Xs(odemodel, forcings = myforcings, events = myevents)}. The definitions of both, forcings and events, are compatible with the way they are defined in \pkg{deSolve}.

@@

It is one of the fundamental concepts of \pkg{dMod} to allow concatenation of functions via the \code{"*"} operator. The mathematical formulation $y = g\circ x$, i.e.~$y(t, p) = (g\circ x)(t, p):= g(x(t, p), p, t)$ becomes \code{y = g*x} in \proglang{R}. To obtain derivatives $\frac{\partial y}{\partial p}$, the chain rule is applied, yielding $\frac{\partial y}{\partial p} = \frac{\partial g}{\partial x}\frac{\partial x}{\partial p} + \frac{\partial g}{\partial p}$. This means that \code{Y()} needs to be informed which symbols are states and parameters to generate the corresponding expressions $\frac{\partial g}{\partial x}$ and $\frac{\partial g}{\partial p}$. The observation function \code{g} takes care of computing theses derivatives from the symbolic expressions and doing the matrix multiplication with the sensitivities $\frac{\partial x}{\partial p}$ from the prediction function. Evaluation of symbolic expressions can become inefficient in \proglang{R}. Therefore, the observation function is usually translated into a \proglang{C} code and compiled.

@@

A set of parameter values, forcings and events capture a certain condition in which we find the modeled system. Manipulating the system, single model parameters, forcings or events need to be changed to account for the manipulation. It is a typical approach to systems inference to systematically perturb small parts of a modeled system to reveal information about the processes.

@@

Prediction- and observation functions are generalized to multiple conditions by the \code{"+"} operator, too. Mathematically speaking, they become
\begin{align*}
	\begin{aligned}
		x: &  & \mathbb R \times \bigoplus_{i = 1}^n \mathbb R^q & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^m \\
		 & & (t, p_1, \dots, p_n) & \longmapsto (x_1(t, p_1), \dots, x_n(t, p_n)) \\
		g: &  & \bigoplus_{i = 1}^n \mathbb R^m \times \bigoplus_{i = 1}^n \mathbb R^q\times \mathbb R& \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^s \\
		& & (x_1(t), \dots, x_n(t), p_1, \dots, p_n, t) & \longmapsto (g_1(x_1(t), p_1, t), \dots, g_n(x_n(t, p_n), p_n, t)).
	\end{aligned}
	\label{}
\end{align*}
In words, if prediction- or observation functions are defined for different conditions then they expect condition-specific inputs which are evaluated by the matching functions. Examples for condition-specific prediction functions typically involve different forcings or events. Observation functions can e.g.~differ betwee different measurement techniques. In all these cases, the different prediction functions $x_1, \dots, x_n$ or observation functions $g_1, \dots, g_n$ are defined, referencing the condition, and combined by the \code{"+"} operator analogously to \code{p}.
% Aufruf der Funktionen mit subset von conditions

All commands, \code{P()}, \code{Xs()}, \code{Y()}, etc.~can be executed with \code{condition = NULL}. In that case, the corresponding returned function is generic and, if called for different conditions, the same identical function is evaluated with the condition-specific input. The other way round, if a function say \code{x} is defined for several conditions, its prediction can be evaluated only for a subset of conditions by \code{x(times, pars, conditions = myconditions)}.

@@

Objective functions can be added by the \code{"+"} operator, meaning the objective values, gradients and Hessians are accumulated by summation. Besides the standard function \code{normL2()}, the functions \code{constraintL2()} and \code{datapointL2()} produce objective functions, too. They allow to define parameter priors or to treat data points as parameters. Thus, a typical objective function used for parameter estimation could be
\begin{center}
	\code{obj <- normL2(data, g*x*(p1+p2+p3)) + constraintL2(mu, sigma)}.
\end{center}

@@

The following model is a simplified version of the dynamic model on bile acid transport published in [REF]. Bile acids are produced in the liver. They are necessary for the digestion of fat and oil. In the liver, bile acids are taken up in hepatocytes (liver cells) by specific transporter molecules. Clearance occurs either via canalicular or sinusoidal export, i.e.~export to the bile transportation system of bile canaliculi or the intercellular space.

>>>>

%% an abstract and keywords
\Abstract{
In a wide variety of research fields, dynamic modeling is employed as an instrument to learn and understand complex systems. The differential equations involved in this process are usually non-linear and depend on many parameters whose values decide upon the characteristics of the emergent system. The inverse problem, i.e.~the inference or estimation of parameter values from observed data, is of interest from two points of view. First, the existence point of view, dealing with the question whether the system is able to reproduce the observed dynamics for any parameter values. Second, the identifiability point of view, investigating invariance of the prediction under change of parameter values, as well as the quantification of parameter uncertainty.

@@

The key elements of the framework and the methodology implemented in \pkg{dMod} are highlighted by an application on a three-compartment transporter model.

@@

}
\Keywords{Dynamic models, parameter estimation, code generation, maximum likelihood, uncertainty analysis}
\Plainkeywords{Dynamic models, parameter estimation, code generation, maximum likelihood, uncertainty analysis} %% without formatting
%% at least one keyword must be supplied

@@

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
\section{Introduction}
Dynamic models are found in several research fields, such as physics, biology or finance. In all these fields, models link theoretical concepts and empirical evidence.
%In a dynamic model, the single terms correspond to mechanisms or processes which are combined to a whole, giving rise to a complex system. If the single terms are associated to parameters, parameter estimation can identify those processes which are crucial to explain the observation.
Single mechanisms or single processes of a complex system are represented by respective terms in the equations of a dynamic model. Parameter estimation can then identify those processes which are crucial to explain the observation.
In that sense, parameter estimation can be employed as an instrument to \textit{understand} complex systems. Once, the link between observation and model is established by the estimated parameters, questions about their identifiability arise. The parameter space needs to be explored to analyze whether the estimate is unique and to determine confidence bounds.

@@

Another distinguishing feature of \pkg{dMod} is the handling of non-identifiability of parameters, a phenomenon occurring frequently in the context of parameter estimation in dynamic systems.
In some cases, non-identifiability has structural reasons. The differential equations bear certain symmetries and these can or cannot be broken, depending on the structure of the observation. A functional relationship between parameters that leaves the observation invariant is the consequence of the latter. In other cases, the data allows to determine a unique optimum but other solutions, although worse, cannot be statistically rejected. The \pkg{dMod} package deals with parameter identifiability and parameter uncertainty by the profile-likelihood method \citep{murphy2000profile, raue2009structural, kreutz2013profile}. The method has proven especially useful in the case of non-identifiable parameters where results obtained from both, the quadratic approximation by the variance-covariance matrix and MCMC sampling can be misleading \citep{raue2013joining}. Besides parameter uncertainties, the profile likelihood method allows to estimate prediction uncertainty \citep{kreutz2012likelihood}. It therefore supports the planning of new informative experiments \citep{raue2011addressing} or suggests possible model reductions \citep{maiwald2016driving}.

@@

\subsection{Dynamic models and model sensitivities}
Dynamic models describe systems with states $x$, usually quantifying the involved species, their interaction and evolution over time. %  as e.g. given by chemical reaction networks.
The time evolution of the states is expressed via a set of ODEs, $\dot x = f(x)$. Although constituting quite a special class of dynamic systems, chemical reaction networks formulated by the law of mass action as considered here allow for surprisingly general applications. Typical ODE examples derived from the law of mass action are:
\begin{itemize}
	\item $\dot x_A = k_p$, constant production of $x_A$ ($\emptyset \rightarrow A$)
	\item $\dot x_C = k_c \cdot x_A \cdot x_B = - \dot x_A = - \dot x_B$, complex formation ($A+B \rightarrow C$)
	\item $\dot x_C = - k_d \cdot x_C$, proportional degradation of $x_C$ ($C \rightarrow \emptyset$)
\end{itemize}
each generally written as
\begin{align}
	\dot x = \underbrace{k_0 \left(\prod x_i^{\nu_i}\right)}_{v_0(x, k_0)} (\nu' - \nu) \quad\leftrightarrow\quad (\nu_1 X_1 + \dots + \nu_n X_n \longrightarrow \nu'_1 X_1 + \dots + \nu'_n X_n)
	\label{eq:ma}
\end{align}
where $X_1, \dots, X_n$ are the $n$ species involved, $\nu$ and $\nu'$ encode the stoichiometry of the reaction and $k_0$ is the reaction rate. When several reactions are involved, eq.~\eqref{eq:ma} generalizes to
\begin{align}
	\dot x = S\cdot v(x, k)
	\label{eq:Sv}
\end{align}
%$\dot x_l = \sum_{j = 1\dots m}  k_{j} \prod_{i = 1\dots n} \nu_{i,j} x_i^{\nu_{i,j}}$ for a system of $n$ states and $m$ reactions with rate constants $p = \{ k_j, j = 1\dots m\}$ and stoichiometric constants $\nu_{i,j} \in \mathbb{Z}$.
%An even more general expression of the system is given by $\dot x = S v(x,p)$
with the stoichiometry matrix $S$
%containing the stoichiometric constants $\nu_{i,j}$
obtained from the collection of vectors $\nu' - \nu$ for reactions $1, \dots, r$,
and the flux vector $v(x,k)\in\mathbb R^{r}$.
Eq.~\eqref{eq:Sv} also holds for general ODEs that do not follow from mass action kinetics. Moreover the system may be explicitly time-dependent and may contain forcings $u(t)$ which e.g.~describe the external stimulation of the system. A general form of the dynamic model is therefore given by
\begin{align}
	\dot x = f(x,k,u,t) = S \cdot v(x,k,u,t), \quad\quad \textrm{with }  x(t = 0) = x_0.
\end{align}
The solution $x(t,\theta)$ for a given parameter vector $\theta = (k, x_0)$ is called model prediction. %The parameters $\theta$ contain the rate constants, initial values $x_0$ and additional parameters included in the forcings.
It is generally assumed that the stoichiometry matrix $S$ of the system is known.
Besides the model predition itself, also the sensitivity of the prediction to changes in the parameter values is of interest.
The sensitivities $s_i = \frac{\partial x}{\partial \theta_i}$ satisfy the \textit{sensitivity equations}
\begin{align}
	\dot s_i  = \frac{\partial f}{\partial x} s_i + \frac{\partial f}{\partial \theta_i}, \qquad \textrm{with }s_i(t = 0) = \frac{\partial x_0}{\partial \theta_i},
\end{align}
a system of ordinary differential equations that in general depend on the states $x$ and, therefore, need to be solved jointly with the original ODE.

Often, the experimentally observed quantities $y$ do not directly correspond to the species $x$ described by the ODE, but are obtained via an observation function $g: \mathbb{R}^n \rightarrow \mathbb{R}^m$,  \begin{align}
	y = g(x, c)
	\label{eq:observation}
\end{align}
with the observation parameters $c$. Examples for observation functions are scaling and offset transformations, $y = c_s \cdot x + c_o$, or the measurement of a superposition of species, $y_i = \sum_j c_j x_j$. The model prediction for the observed states is obtained by evaluating the observation function on the solution of the ODE. Following the chain rule of differentation also the sensitivities of the observed states $\frac{\partial y}{\partial \theta_i} = \frac{\partial g}{\partial x} \frac{\partial x}{\partial\theta_i}  + \frac{\partial g}{\partial \theta_i}$ are obtained. Here, the parameter vector $\theta$ has been augmented by the observation parameters, $\theta = (k, x_0, c)$.

@@

The easiest and one of the most frequent situations is when $\phi$ follows a normal distribution, $\phi\sim\mathcal N(y(\theta), \Sigma)$. When measurements are statistically independent, the variance-covariance matrix $\Sigma = {\rm diag}(\sigma_1^2, \dots, \sigma_N^2)$ is diagonal and $\phi$ factorizes. The likelihood function reads 
\begin{align}
	L(\theta) &= \prod_{i = 1}^N \frac{1}{\sqrt{2\pi\sigma_i^2}} e^{-\frac{(y_i(\theta)-y_i^D)^2}{2\sigma_i^2}} .
	\label{eq:likelihood}
\end{align}
Taking twice the negative logarithm, eq.~\eqref{eq:likelihood} turns into
\begin{align}
	l(\theta) = -2\log L(\theta) &= \sum_i \left[\left(\frac{y_i(\theta) - y_i^D}{\sigma_i}\right)^2 + \log(2\pi\sigma_i^2)\right],
	\label{eq:nll}
\end{align}
thus converting the maximization of $L(\theta)$ to a minimization of $l(\theta)$.
Assuming that the data uncertainties $\sigma_i$ are known, eq.~\eqref{eq:nll} is the weighted least-squares function shifted by a constant. For unknown $\sigma_i$, eq.~\eqref{eq:nll} can be optimized with respect to $\theta$ and $\sigma$ jointly, yielding the MLE $\hat\theta^{\prime} = (\hat\theta, \hat\sigma)$.

@@

Optimization by the Newton method attempts to iteratively find the root of the gradient $\nabla l (\theta)$ by the recursion
\begin{align}
	\theta^{(n+1)} = \theta^{(n)} - Hl(\theta)^{-1}\nabla l(\theta),
	\label{eq:newton}
\end{align}
where $Hl(\theta)$ denotes the Hessian, i.e.~the matrix of second derivatives of $l(\theta)$. The fact that for normally distributed noise the log-likelihood is a least squares function, has a big advantage: gradient and (approximate) Hessian can be computed from first-order derivatives  \citep{press1996numerical}. Since the second-order contributions to the Hessian can be neglected, gradient and Hessian read
\begin{align}
	\nabla l(\theta) = 2 r(\theta) J(\theta) \quad\quad Hl(\theta) \approx 2 J(\theta)^T J(\theta),
	\label{eq:derivatives}
\end{align}
with the weighted residual vector $r_i(\theta) = \frac{y_i(\theta) - y_i^D}{\sigma_i}$ and its first derivative, the Jacobian matrix
\begin{align}
	J_{ij}(\theta) = \frac{\partial r_i}{\partial \theta_j}(\theta) = \frac{1}{\sigma_i} \frac{\partial y_i}{\partial \theta_j}(\theta). \label{eq:jacobian}
\end{align}
As indicated by eqs.~\eqref{eq:derivatives}-\eqref{eq:jacobian},  first order model sensitivities $\frac{\partial y_i}{\partial\theta_j}$ are sufficient to compute gradient and Hessian in good approximation.

@@

Parameter estimates $\hat \theta$  are obtained by non-linear optimization of the log-likelihood function $l(\theta)$. Although being a function of the parameters,  the log-likelihood depends on the data, too. Consequently, "equivalent" random realizations of the data would lead to different parameter estimates $\hat\theta$. Given $\hat\theta$ for one random realization of the data, the question is, at which significance level we can reject other parameters $\theta$. This leads to the related question of parameter convidence intervals.

A useful tool to derive confidence intervals beyond the scope of Fisher Information Matrix is the profile likelihood. Consider a parameter of interest, $\theta_i$, being one of the parameters in the vector $\theta$. Furthermore, let $c_{\tau}(\theta) = \theta_i - \tau = 0$ be a parameter constraint.
Then, extending the likelihood with the constraint via a Lagrange multiplier $\lambda$ yields
\begin{align}
	\begin{aligned}
		{\rm pl}_i:\quad \mathbb R & \longrightarrow \mathbb R\\
		\tau & \longmapsto \min_{(\theta, \lambda)} \big[ l(\theta) + \lambda c_{\tau}(\theta) \big]
	\end{aligned}
	\label{eq:pl}
\end{align}
which is called the profile likelihood of the $i^{\;\rm th}$ parameter. The path $\tau \mapsto (\hat\theta_{\tau}, \hat\lambda_{\tau})$ along the minimizing parameters is called the profile likelihood path. Hence, the profile likelihood ${\rm pl}_i$ returns the minimal log-likelihood under the constraint $\theta_i \stackrel{!}{=} \tau$. By construction, the inequality $D_i := {\rm pl}_i(\tau) - {\rm pl}_i(\hat \theta_i) \geq 0$ holds for all $\tau$, assuming equality at least for $\tau = \hat\theta_i$. Interpreting the unconstrained model as the null model, which we assume to be true, and that with $\theta_i$ fixed to $\tau$ as the alternative model, the value $D$ is twice the log of the likelihood-ratio between those. Hence, a likelihood-ratio test can be performed to accept or reject the alternative model at a given confidence level. To compute the corresponding threshold, we assume Wilks' theorem according to which the thresholds are the quantiles of the $\chi^2$ distribution with one degree of freedom. Consequently, the $68\%/90\%/95\%$-confidence intervals of $\theta_i$ are those values of $\tau$ for which $D_i(\tau) \leq 1$, $2.71$ and $3.84$, respectively.

@@

\pkg{dMod} makes use of derivatives wherever possible. It is one of the core functionalities of the \pkg{cOde} package upon which \pkg{dMod} is based to augment a system of ODEs by its sensitivity equations. If $r$ parameters are involved and the system consists of $n$ states, the number of equations grows as quick as $n^2 + nr$. Solving the equations can be considerably accelerated by utilizing compiled code. \pkg{dMod} provides the \code{odemodel} class. An object of class \code{odemodel} is generated from an \code{eqnlist} or \code{eqnvec}. In the background, \proglang{C} code for the ODE and the combined system of ODE and sensitivity equations is generated, written to the working directory and compiled. The \code{odemodel} object keeps track of the shared objects. It is the basis of a prediction function $x(t, p)$ generated by the \code{Xs()} command.

@@

\pkg{dMod} seeks to stay close to the mathematical formulation, i.e.~\code{x <- Xs(myodemodel)} will indeed return an \proglang{R} function, an object of class \code{prdfn}, which expects arguments \code{times} and \code{pars} and turns them into a model prediction. The letter \code{"s"} in \code{"Xs"} refers to sensitivities, i.e.~the solution $x(t)$ is returned, the sensitivities $\frac{\partial x}{\partial p}(t)$ are returned, too.

Besides parameters, the prediction might also depend on forcings $u(t)$ and sudden events, e.g.~setting states to 0 at a predefined point in time. Since forcings and events are by default fixed, they are defined together with the  prediction function, \code{x <- Xs(odemodel, forcings = myforcings, events = myevents)}. The definitions of both, forcings and events, are compatible with the way they are defined in \pkg{deSolve}.

@@

It is one of the fundamental concepts of \pkg{dMod} to allow concatenation of functions via the \code{"*"} operator. The mathematical formulation $y = g\circ x$, i.e.~$y(t, p) = (g\circ x)(t, p):= g(x(t, p), p, t)$ becomes \code{y = g*x} in \proglang{R}. To obtain derivatives $\frac{\partial y}{\partial p}$, the chain rule is applied: $\frac{\partial y}{\partial p} = \frac{\partial g}{\partial x}\frac{\partial x}{\partial p} + \frac{\partial g}{\partial p}$. This means that \code{Y()} needs to be informed which symbols are states and parameters to generate the corresponding expressions $\frac{\partial g}{\partial x}$ and $\frac{\partial g}{\partial p}$. The observation function \code{g} takes care of computing theses derivatives from the symbolic expressions and doing the matrix multiplication with the sensitivities $\frac{\partial x}{\partial p}$ from the prediction function. Evaluation of symbolic expressions can become inefficient in \proglang{R}. Therefore, the observation function is usually translated into a \proglang{C} code and compiled.

@@

A set of parameter values, forcings and events capture a certain condition in which we find the modeled system. Manipulating the system, single model parameters, forcings or events need to be changed to account for the manipulation. It is a typical approach in systems inference to systematically perturb small parts of a modeled system to reveal information about the processes.

@@

Prediction- and observation functions are generalized to multiple conditions by the \code{"+"} operator, too. Mathematically speaking, they become
\begin{align*}
	\begin{aligned}
		x: &  & \mathbb R \times \bigoplus_{i = 1}^n \mathbb R^q & \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^m \\
		 & & (t, p_1, \dots, p_n) & \longmapsto (x_1(t, p_1), \dots, x_n(t, p_n)) \\
		g: &  & \bigoplus_{i = 1}^n \mathbb R^m \times \bigoplus_{i = 1}^n \mathbb R^q\times \mathbb R& \longrightarrow  \bigoplus_{i = 1}^n \mathbb R^s \\
		& & (x_1(t), \dots, x_n(t), p_1, \dots, p_n, t) & \longmapsto (g_1(x_1(t), p_1, t), \dots, g_n(x_n(t, p_n), p_n, t)).
	\end{aligned}
	\label{}
\end{align*}
In words, if prediction- or observation functions are defined for different conditions then they expect condition-specific inputs which are evaluated by the matching functions. Examples for condition-specific prediction functions typically involve different forcings or events. Observation functions can e.g.~differ between different measurement techniques. In all these cases, the different prediction functions $x_1, \dots, x_n$ or observation functions $g_1, \dots, g_n$ are defined, referencing the condition, and combined by the \code{"+"} operator analogously to \code{p}.
% Aufruf der Funktionen mit subset von conditions

All commands, \code{P()}, \code{Xs()}, \code{Y()}, etc.~can be executed with \code{condition = NULL}. In that case, the corresponding returned function is generic and, if called for different conditions, the same identical function is evaluated with the condition-specific input. The other way round, if a function, say, \code{x} is defined for several conditions, its prediction can be evaluated only for a subset of conditions by \code{x(times, pars, conditions = myconditions)}.

@@

Objective functions can be added by the \code{"+"} operator, meaning the objective values, gradients and Hessians are accumulated by summation. Besides the standard function \code{normL2()}, \pkg{dMod} provides several other functions returning objective functions, e.g.~\code{constraintL2()} or \code{datapointL2()}. They allow to define parameter priors or treat data points as parameters, as shown in Section~\ref{sec:example}. Thus, a typical objective function used for parameter estimation could be
\begin{center}
	\code{obj <- normL2(data, g*x*(p1+p2+p3)) + constraintL2(mu, sigma)}.
\end{center}

@@

The following model is a simplified dynamic model on bile acid transport. Bile acids are produced in the liver. They are necessary for the digestion of fat and oil. In the liver, bile acids are taken up in hepatocytes (liver cells) by specific transporter molecules. Clearance occurs either via canalicular or sinusoidal export, i.e.~export to the bile transportation system of bile canaliculi or the intercellular space.

To study bile clearance, experiments with hepatocytes or hepatocyte-derived cell lines are performed \textit{in-vitro}. Cells in a Petri dish stick together to a monolayer of cells, forming bile canaliculi in between cells. They adhere to the dish and are surrounded by a buffer providing the cells with nutrients. A radioactive label allows to measure the bile acid \textit{taurocholic acide} (TCA).
